---
title: "Fldgen Tutorial"
author: "Robert Link"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width=8,
  cache=FALSE,
  comment = "#>"
)
library('fldgen')
library('dplyr')
library('ggplot2')
library('reshape2')
```

The fldgen package allows you to ingest temperature output from an earth system
model (ESM) and generate randomized temperature fields that have the same space
and time correlation properties as the original ESM data.  This tutorial focuses
on how to use the functions in the package to generate and analyze temperature
fields.  The details of how the method works are covered in a companion paper. 

All of the functions used here are documented in R's help system.  Since our
purpose here is to outline what functions have to be called, and in what
sequence, to perform the analysis, we haven't repeated material from the help
files.  If you're confused about how a function is supposed to work, consult the
help files.  For example, `help(read.temperatures)` will print the docs for the function
that reads the netCDF temperature fields.

## Using the field generator
### Reading the ESM data
```{r params}
## parameters for the code below.
ngen <- 4             # number of fields to generate
exfld <- 20           # example field to plot from the time series
set.seed(8675309)     # Set RNG seed so results will be reproducible
```

All of the data needed for this tutorial is installed with the package.
```{r datadir}
library('fldgen')
datadir <- system.file('extdata', package='fldgen')
```

The ESM temperature field should be in a netCDF file.  The function that reads
it returns a "griddata" structure that contains the data, as well as some
information about the grid, such as the latitude, longitude, and time
coordinates and the vector of area weights needed to compute grid averages.

```{r readdata}
griddata <- read.temperatures(file.path(datadir, 'tann1.nc'))
tgav <- griddata$tas %*% griddata$tgop
```
Here, we have read in the netCDF data and used the global average operator to 
compute a time series of global mean temperatures.  These variables will be used
as input to the functions that analyze the ESM data and produce the temperature
fields.

### Generating fields

The first thing we need is a model for the mean temperature response in each
grid cell.  That is, for each grid cell, what is the mean temperature of that
cell as a function of global mean temperature.  You can use whatever model you
want for this.  We will use a simple linear pattern scaling model, which is
implemented in the `pscl_analyze` function.
```{r pscl}
pscl <- pscl_analyze(griddata$tas, tgav)
```

The mean response analysis should return the model coefficients (called `w` and
`b` in our linear model) and a time series of gridded residuals.  These
residuals encode all of the information about the spatial variability of the
ESM, so it is these we will pass to the empirical orthogonal functions (EOF)
analysis.  

```{r eof}
reof <- eof_analyze(pscl$r, griddata$tgop)
```

The last thing we need to generate our fields is the temporal structure of the
EOF coefficients.  We get this from the Fourier transform of the coordinates of
the residuals in the coordinate system defined by the EOF basis vectors.

```{r fft}
Fx <- mvfft(reof$x)
Fxmag <- abs(Fx)
Fxphase <- atan2(Im(Fx), Re(Fx))
```

Now we're ready to generate fields.  We'll do `r ngen` fields in this example. 
The first one will be a reconstruction of the original input.  The other three
will new fields.

```{r genfields}
tempgrids <- list()
length(tempgrids) <- ngen

## Here tgav is the same as the input, but that doesn't have to be the case
meanfield <- pscl_apply(pscl, tgav)

## First field will have the same phases as the input ESM data
tempgrids[[1]] <- reconst_fields(reof$rotation, mkcorrts(Fxmag, Fxphase), meanfield)

## Other fields will have random phases
for(i in 2:ngen) {
  tempgrids[[i]] <- reconst_fields(reof$rotation, mkcorrts(Fxmag), meanfield)
}

## Subtract off the mean field.  Save these because we will want to use them later
residgrids <- lapply(tempgrids, function(g) {
    g - meanfield
})

```

### Plotting the fields

We can extract a single field from each time series and plot them all for
comparision.  We will be able to see a lot more detail if we subtract out the
mean field from each one, so that's what we will do.

```{r plotfields, fig.show='hold', eval=TRUE}
## Extract a single example field from each series and create a plot
fieldplots <- lapply(residgrids, function(g) {
    suppressWarnings(
        plot_field(g[exfld,], griddata, 14, titlestr='delta-T (K)')
    )
})

## Display the plots
for (plt in fieldplots) {
    print(plt)
}
```

## EOF Characteristics

### Time behavior

Start by constructing a heatmap of the power spectrum for the EOFs.
Technically we will be plotting the square root of the power spectrum,
which is fine for getting a sense of what trends exist in the data,
but we'll need to keep that in mind if we decide to do anything
quantitative with the power.

```{r heatmap}
nt <- length(griddata$time)
## There is no additional information in the negative frequencies, so keep only
## the positive ones.
np <- 
    if(nt %% 2 == 1) { 
        (nt+1)/2
    } else {
        nt/2 + 1
    }

hmdata <- reshape2::melt(Fxmag[1:np,])
hmdata$freq <- (hmdata$Var1 - 1) / nt    # The - 1 is due to R's unit-indexed arrays.
hmdata$EOF <- as.integer(substring(hmdata$Var2, 3))
## discretize the power so we can isolate structure more easily
nbrk <- 10
hmdata$discval <- findInterval(hmdata$value / max(hmdata$value), seq(0.01, 0.99, length.out=nbrk)) / nbrk
hmdata <- dplyr::select(hmdata, EOF, freq, value, discval) %>% as_tibble()

hmplt <- ggplot(hmdata, aes(x=freq, y=EOF, fill=discval)) + geom_raster() + scale_fill_distiller(palette='YlOrRd', direction=1, name='sqrt(Relative Power)')

print(hmplt)
```

Evidently, the power drops off quite a bit after the first few EOFs.
```{r eofpwr}
eofpwr <- group_by(hmdata, EOF) %>% summarise(totpwr=sum(value*value)) %>% mutate(relpwr=totpwr/max(totpwr)) %>% select(EOF, relpwr)
eofpwrplt <- ggplot(eofpwr, aes(x=EOF, y=relpwr)) + geom_col() + ylab('Relative Power')
print(eofpwrplt)
```
More importantly, the power spectrum whitens after the first few EOFs, so those
early EOFs represent periodic signals, while the later ones don't.  Here are the
smoothed power spectra for the first 9 EOFs.

```{r pwrspectra}
eofspectra <- filter(hmdata, EOF>0, EOF<10) %>% mutate(EOF=factor(EOF))
eofspectraplt <- ggplot(eofspectra, aes(x=freq, y=value, color=EOF)) + geom_smooth(se=FALSE) + scale_color_brewer(palette='Set1') + ylab('freq (1/yr)')
print(eofspectraplt)
```

### Spatial behavior

We will make spatial plots of the 9 EOFs shown in the plot above.  We'll also
plot EOFs 25 and 50, just to get an idea of what's happening in those lower
power modes.

```{r eofspatial}
### Plotting global maps is still a little slow, so expect this to take some time.
## The EOFs are in reof$rotation.  Each column is the grid cell values for an EOF.
## Also, the EOFs are scaled to unit norm.  We'll rescale them to unit max value

eofcols <- c(2:10, 26, 51)  # EOF numbering starts at 0, but array numbering starts at 1
eofvis <- t(reof$rotation[,eofcols])   # EOFs are now in rows, not columns
eofvis <- eofvis / max(abs(eofvis))

eofplts <- lapply(seq_along(eofcols), function(i) {
    suppressWarnings(plot_field(eofvis[i,], griddata, 14, -0.9, 0.9, titlestr='delta-T (K)'))
})
    
for(i in seq_along(eofcols)) {
    title <- paste0('EOF-', eofcols[i]-1)
    print(eofplts[[i]] + ggtitle(title))
}
```
